---
output: github_document
---

# Creating ENVI files for Hypertrace

Hypertrace is only set up to work with ENVI binary files.
These files can be easily created using GDAL utilities.
This wiki page shows some examples for creating ENVI files in R and Python.

## R

We'll start by generating an example set of spectra from simulations using the PRO4SAIL canopy radiative transfer model, as implemented in the [`rrtm` package](https://github.com/ashiklom/rrtm).
However, any similarly-formatted set of spectral data will work.

```{r}
# install.packages("remotes")
# remotes::install_github("ashiklom/rrtm")
library(rrtm)
laiseq <- seq(0.5, 5, 0.5)
rawspec <- lapply(laiseq, pro4sail_5, N = 1.4, Cab = 40, Car = 8,
                  Cbrown = 0, Cw = 0.01, Cm = 0.01, soil_moisture = 0.8)
refl <- lapply(rawspec, "[[", "hdr")
wavelengths <- seq(400, 2500)
spec_matrix <- do.call(cbind, refl)
dim(spec_matrix)
head(spec_matrix)
```

Each row here is a wavelength (400-2500 nm), and each column is a spectrum.

```{r}
colors <- hcl.colors(length(laiseq), palette = "viridis") 
matplot(wavelengths, spec_matrix, type = "l", lty = "solid",
        xlab = "Wavelength (nm)", ylab = "Reflectance [0,1]",
        col = colors)
legend("topright", as.character(laiseq), lty = "solid",
       col = colors, lwd = 2, title = "LAI")
```

We will use the R `raster` (available on CRAN) to convert this matrix to a raster.
The `raster` package expects these data to be a 3D array with the first two dimensions corresponding to space and the last dimension containing the bands (i.e., a "band-sequential", or "BSQ" interleave format).
We can easily convert our spectra matrix to this format using base R functions.

```{r}
# First, convert the matrix to an array.
spec_array <- array(spec_matrix, c(dim(spec_matrix), 1))
dim(spec_array)
# Then, reshape the array to match what raster expects. Here, we use aperm to
# reverse the dimensions (3rd dimensions first, 2nd dimensions second, 1st
# dimension last).
spec_array_t <- aperm(spec_array, c(3, 2, 1))
dim(spec_array_t)
```

Now, we can create a `raster::brick` object from this array.

```{r}
# install.packages("raster")
library(raster)
spec_raster <- brick(spec_array_t)
print(spec_raster)
```

Finally, we can write this object to disk, specifying the output format and interleave.
(NOTE: Isofit currently only accepts BIP and BIL interleaves. Its default output is BIP, so that's what we will use here.)

```{r}
writeRaster(spec_raster, "my-raster-file.envi", format = "ENVI",
            options = c("INTERLEAVE=BIP", "SUFFIX=ADD"))
```

We're not quite done because we need to add additional metadata about the wavelengths to the `my-raster-file.envi.hdr` header file.
We can use base R string processing functions to create the corresponding strings and add them to the bottom of the header file.

```{r}
hdrfile <- "my-raster-file.envi.hdr"
write("wavelength units = Nanometers", hdrfile, append = TRUE)
wavelength_string <- paste0("wavelength = {", paste(wavelengths, collapse = " , "), "}")
write(wavelength_string, hdrfile, append = TRUE)
# Full-width-half-max, in nm. Ours is 1nm.
fwhm <- rep(1.0, length(wavelengths))
fwhm_string <- paste0("fwhm = {", paste(fwhm, collapse = " , "), "}")
write(fwhm_string, hdrfile, append = TRUE)
```

Let's confirm that this worked by trying to read in the file we just created.

```{r}
rast <- brick("my-raster-file.envi")
rast_matrix <- drop(rast[])
dim(rast_matrix)
matplot(wavelengths, t(rast_matrix), type = "l",
        xlab = "Wavelength (nm)", ylab = "Reflectance [0,1]")
```

## Python

We can follow a similar procedure using Python and the `spectral` module.
We'll start with the same dataset we just created, but first convert it to a `numpy` array to simulate working with real data.

```{python}
# pip install spectral numpy
import spectral as sp
import numpy as np

img = sp.open_image("my-raster-file.envi.hdr")
spec_array = img.asarray()
print(spec_array.shape)

spec_matrix = np.squeeze(spec_array).transpose()
print(spec_matrix.shape)
```

Now, we have the same spectra matrix that we used with R.
Let's also define the image wavelength and FWHM.

```{python}
wavelengths = np.linspace(400, 2500, 2101)
fwhm = np.ones(len(wavelengths))
```

Next, similarly to the R workflow, we convert the matrix into an array with dimensions lines x samples x bands.

```{python}
# Add a length-1 dimension to the end
output_array = np.expand_dims(spec_matrix, 2)
output_array.shape
output_array_t = np.transpose(output_array, (2, 1, 0))
output_array_t.shape
```

Finally, we can use the `spectral.envi.save_image` function to save this array.
Note that the wavelengths and FWHM are stored in the `metadata` `dict`, which can also be used to store any other header items (e.g., file description, spatial information).

```{python}
metadata = {"wavelength": wavelengths, "fwhm": fwhm}
out_raster = sp.envi.save_image("my-python-raster.hdr", output_array_t,
                                ext=None, interleave="bip",
                                metadata=metadata)
```

Finally, let's confirm that we can open this image.

```{python}
test = sp.open_image("my-python-raster.hdr")
print(test)
test_m = test.open_memmap()
test_m.shape
print(np.all(test_m == output_array_t))
```
